% gwebman.tex -- the GWEB manual.  Process with a TeX engine that can find
% gwebmac.tex, for example
%       TEXINPUTS=../tex: pdftex gwebman.tex
% or simply `make manual' from the top of the repository.
%
% It is formatted in the manner of Knuth and Levy's cwebman.tex: it loads the
% GWEB macro package (the same one gweave's output uses) and adds its own
% section, list, verbatim, and running-head macros on top.

\input gwebmac
\Gpdffalse \Gdvipdffalse        % a plain manual: no hyperlinks or bookmarks
\parskip=0pt plus 1pt
\parindent=18pt
\abovedisplayskip=.5\abovedisplayskip  \belowdisplayskip=.5\belowdisplayskip
\abovedisplayshortskip=.5\abovedisplayshortskip
\belowdisplayshortskip=.5\belowdisplayshortskip

\font\eighttt=cmtt8
\font\ninett=cmtt9
\font\twelvett=cmtt12
\font\authorfont=cmr12
\font\sectionfont=cmbx12

% --- inline code; @ < > = etc. are safe in \tt, so avoid braces inside \cw ---
\def\cw#1{{\tt #1}}
\def\bs{\char92 }                              % a literal backslash, in \tt
\def\rn#1{$\langle\,${\it #1\/}$\,\rangle$}    % a refinement name, as in print

% --- a centred chapter heading, in the manner of cwebman's \section ---
\def\chapter#1{\penalty-500\bigskip
  \centerline{\sectionfont\def\.##1{{\twelvett##1}}#1}\nobreak\vskip 6pt}
% --- a run-in bold subsection heading ---
\def\section#1{\medskip\noindent{\bf #1.}\enspace\ignorespaces}

% --- a bulleted list item ---
\def\bull{\smallskip\item{$\bullet$}}

% --- a verbatim code display (TeXbook recipe): | is the escape inside it ---
\def\begincode{\par\medskip\begingroup
  \def\do##1{\catcode`##1=12 }\dospecials
  \catcode`\|=0
  \obeylines \obeyspaces \tt \parindent=0pt \leftskip=\parindent\relax}
\def\endcode{\endgroup\medskip\noindent}

% --- running heads, in the manner of cwebman ---
\def\runninghead{{\ninett GWEB} USER MANUAL (VERSION 0.4.0)}
\def\lheader{\mainfont\folio\hfil{\sc\runninghead}\hfil}
\def\rheader{\hfil{\sc\runninghead}\hfil\mainfont\folio}

% =============================================================================
% Title page (no running head), in the manner of cwebman.
\headline={\hfil}\footline={\hfil}
\null\vfill
\centerline{\titlefont The {\ttitlefont GWEB} System}
\vskip 6pt
\centerline{\sectionfont A Literate Programming System for Go}
\vskip 18pt
\centerline{(Version 0.4.0)}
\vskip 24pt
\centerline{\authorfont modeled on Knuth and Levy's {\ttitlefont CWEB}}
\vfill
\noindent
This is a guide to \.{GWEB} for readers who already know \.{CWEB}, or
literate programming generally. It does not re-explain what literate programming
is; it concentrates on how \.{GWEB} differs from \.{CWEB} and on the features and
commands you will actually use. To a first approximation \.{GWEB} is ``\.{CWEB} with \CEE/
replaced by \GO/,'' so most of your \.{CWEB} knowledge transfers intact --- the
interesting parts are where \GO/, and \cw{gofmt}, change the picture.
\vfil\eject

% Running heads from the second page on.
\headline={\ifodd\pageno\rheader\else\lheader\fi}

\chapter{\.{GWEB} at a glance}
\noindent\.{GWEB} mirrors \.{CWEB} tool for tool:

\bull \cw{gtangle} $\approx$ \cw{ctangle} --- extracts the compilable \GO/ source;

\bull \cw{gweave} $\approx$ \cw{cweave} --- typesets the documentation;

\bull \cw{gwebmac.tex} $\approx$ \cw{cwebmac.tex} --- the \TeX\ macros the woven
output uses.

\noindent A web is a single file, conventionally \cw{foo.w}. \cw{gtangle foo}
writes \cw{foo.go} (plus any extra files); \cw{gweave foo} writes \cw{foo.tex},
which \cw{pdftex} turns into \cw{foo.pdf}. As in \.{CWEB}, the \cw{.w} may be left
off the command line.

\chapter{What carries over from \.{CWEB}}
\noindent
If you know \.{CWEB} you already know most of \.{GWEB}. A web is a sequence of numbered
{\it sections}, each with up to three parts in order --- a \TeX\ commentary
part, a definition part (\cw{@d}/\cw{@f}/\cw{@s}), and a code part --- preceded
by {\it limbo}. The control codes keep their familiar meanings:

\bull \cw{@*}, \cw{@*1}, \cw{@*2}, $\ldots$ start starred (chapter) sections;
\cw{@\ } starts an ordinary section.

\bull \cw{@c} starts unnamed code; \cw{@<Name@>=} defines a refinement and
\cw{@<Name@>} references one. A trailing \cw{...} abbreviates a name to its
unique prefix.

\bull \cw{@(file@>=} sends the following code to another output file; \cw{@i}
includes another web.

\bull \cw{@f}/\cw{@s} format an identifier; \cw{@\char94 text@>}, \cw{@.text@>},
\cw{@:text@>} add index entries; \cw{@!} forces an index definition.

\bull \cw{@=...@>} is verbatim, \cw{@t...@>} is \TeX\ text, \cw{@\char38} pastes
tokens, \cw{@@} is a literal \cw{@}, and \cw{@\char37} and \cw{@q...@>} are
comments.

\bull \cw{@,} \cw{@/} \cw{@|} \cw{@\#} fine-tune the woven layout.

\bull change files (\cw{@x}/\cw{@y}/\cw{@z}) and inline code \cw{|...|} in
commentary work exactly as in \.{CWEB}.

\noindent Two small conveniences: \cw{@p} is accepted as a synonym for \cw{@c}
(as in \.{CWEB}), and you never write \cw{\bs input gwebmac} --- \cw{gweave} emits it
for you, and removes a stray copy if you include one.

\chapter{What makes \.{GWEB} different}
\noindent The differences all follow from typesetting \GO/ instead of \CEE/.

\section{No preprocessor}
\GO/ has no preprocessor, so \cw{@d} never generates code: \cw{gtangle} ignores it
(use \GO/ \cw{const} and \cw{func} for real definitions). It is not useless,
though --- \cw{gweave} reads \cw{@d Name} as a request to set \cw{Name} in
typewriter, like a \.{CWEB} macro (see below). The formatting directives \cw{@f} and
\cw{@s} also remain --- for example \cw{@f Stack int} sets the user type
\cw{Stack} in bold, the way predeclared types are set.

\section{Pretty-printing without a parser}
\cw{cweave} parses \CEE/ and re-derives every space, break, and indent from the
grammar. \cw{gweave} reaches the same math-like result --- a product \cw{a * b}
spaced, a pointer \cw{*T} tight, relations set off, a keyword followed by a space
--- but from {\it local\/} rules only: it looks at each token and its neighbour,
never building a parse tree. The spacing is therefore \cw{cweave}'s, not
\cw{gofmt}'s: where \cw{gofmt} tightens a higher-precedence factor (\cw{a*b + c}),
\cw{gweave} spaces every operator alike, the way a formula is set. Indentation is
likewise structural (see {\it The woven document\/}), so ragged or crammed source
is laid out cleanly all the same --- you need not format your sections by hand.

Two ambiguities have no local answer, and \cw{gweave} leans on the one signal the
author reliably gives, their own spacing, exactly as \cw{cweave} leans on
\cw{@[}\thinspace$\ldots$\thinspace\cw{@]}: a \cw{*} between operands is a pointer
type when the source runs it into what follows (\cw{p *T}) and a product otherwise
(\cw{a * b}, \cw{a*b}); a brace opens a block, spaced, or a composite literal,
tight, as its keyword or type says. \.{CWEB}'s prettyprinter hints \cw{@+},
\cw{@[}\thinspace$\ldots$\thinspace\cw{@]}, and \cw{@;} have nothing to do here and
are ignored; the explicit layout codes \cw{@,} \cw{@/} \cw{@|} \cw{@\#} still
override the automatic spacing for a manual touch-up.

\section{gtangle runs gofmt}
The tangled output is passed through \cw{gofmt}, so a correct web always yields
canonically formatted \GO/. If the assembled text is not valid \GO/, \cw{gtangle}
keeps the unformatted output and prints a warning instead of failing.

\section{Types and constants are classified automatically}
\cw{gweave} scans for \cw{type} declarations --- both the single form (\cw{type
Name ...}) and the parenthesized block --- and sets each declared name {\bf bold}
wherever it appears, like a predeclared type. It also recognizes an \cw{iota}
{\it enumeration}: a \cw{const (...)} block whose first entry seeds the counter
with \cw{iota}. Because these are \GO/'s manifest constants, the closest analogue of
a name defined by \cw{@d}, each is set in typewriter everywhere it appears, the
way \cw{cweave} sets a \cw{@d} macro. A plain \cw{const} --- a one-line \cw{const
Name = ...}, or a block with no \cw{iota} --- is left an ordinary italic
identifier; mark it \cw{@d} yourself if you want it in typewriter. The same goes
for names you do {\it not\/} declare in this web: \cw{@d http.StatusOK} sets
\cw{StatusOK} in typewriter everywhere it appears. You can override any automatic
class with \cw{@f}/\cw{@s} --- e.g.\ \cw{@f Name int} forces \cw{Name} bold, or
\cw{@f Name x} (any ordinary identifier) restores plain italic.

Such a directive may be {\it qualified\/} to a package or receiver. \cw{@s
foo.Bar int} sets only the \cw{Bar} written as \cw{foo.Bar}, so a coincidental
\cw{abc.Bar} (a \cw{Bar} field of some \cw{abc}) --- or a bare \cw{Bar} --- is
left alone; the unqualified \cw{@s Bar int} sets every \cw{Bar}. The qualifier is
just the identifier written before the \cw{.}, so it need not name a real
package.

\section{\.{GWEB} is written in itself}
Like \.{CWEB}, \.{GWEB} is a literate program: each tool is tangled from a \cw{.w} web
kept beside its \GO/. The committed \GO/ is byte-for-byte reproducible from those
sources --- a fixpoint the test suite checks on every change --- so the program
you read and the program that runs cannot drift apart.

\chapter{The woven document}
\noindent
\cw{gweave} produces the same furniture as \cw{cweave}: run-in bold section
headings (with a page break before each major starred section), a two-column
index, a {\it Names of the Sections\/} list, a contents page, and section-number
running heads. A whole section --- its commentary and its code --- is kept
together on one page when it fits, and a code part is never split across a page
boundary unless it is itself taller than a page.

\section{How code is set}
Reserved words and user-defined types are bold, other identifiers italic, string
and rune literals typewriter (in \cw{cmtex10}, as in \.{CWEB}), numbers and comments
roman. A few \GO/-specific touches: the shift operators \cw{<<} and \cw{>>} are set
as the tight symbols $\ll$ and $\gg$; a thin space follows a function name or a
keyword before \cw{(}, so \cw{f (x)} does not jam together; and a comment's
\cw{//} is kerned a little tighter.

A code comment's text is copied through as \TeX, exactly as in \.{CWEB}: \TeX\
mathematics between dollar signs is typeset as mathematics, an inline-code span
\cw{|...|} is set as the \GO/ it stands for, \cw{\bs.\char123...\char125} gives
typewriter, and ordinary control sequences work. The price is the same as
\.{CWEB}'s --- the \TeX\ specials are live, so a literal \cw{\char123}, \cw{\$},
\cw{\%}, \cw{\char38}, or backslash in a comment must be escaped or balanced (write
\cw{\bs\char123}, \cw{\bs\$}, a literal bar as \cw{\bs\char124}), and a stray
\cw{\%} comments out the rest of the line. Keep prose that needs much of this in
the \TeX\ part, not the code.

Neither the indentation nor the spacing is copied from the source. Indentation is
{\it structural\/}: \cw{gweave} tracks the block nesting of each code part ---
braces, parentheses, and brackets, with the special cases \cw{gofmt} makes for
\cw{switch} case bodies and for labels --- and lays every line out accordingly. The
horizontal spacing is likewise derived from the grammar, math-like (see {\it
Pretty-printing without a parser\/}). So a section typed with careless indentation,
or crammed with no spaces at all, still reads cleanly --- much as \cw{cweave}
reindents and re-spaces from the \CEE/ grammar.

\section{Index and formatting codes}
The hand-index codes \cw{@\char94 text@>} (roman), \cw{@.text@>} (typewriter),
and \cw{@:text@>} (formatted by \cw{\bs9}) work as in \.{CWEB}, as do \cw{@!} to
force a definition entry and \cw{@f}/\cw{@s} to reclassify a name. An identifier
is marked as ``defined'' (its index number underlined) when it follows \cw{func},
\cw{var}, \cw{const}, or \cw{type}, or sits just left of \cw{:=}; this is a
heuristic, not a type analysis, and \cw{@!} overrides it.

\section{Hyperlinks and bookmarks}
Section numbers shown as references, in the index, in cross-reference notes, and
on the contents page become links, and the starred sections form a PDF outline.
Two back ends produce them, exactly as in \.{CWEB}:

\bull with \cw{pdftex} in PDF mode, the engine's own primitives;

\bull through DVI, by requesting the \cw{\bs special} back end with
\cw{\bs let\bs pdf+} and converting the result with \cw{dvipdfmx}.

\noindent A plain DVI run with no \cw{\bs pdf} simply omits the links and outline.

\section{Helper macros for commentary}
\cw{gwebmac.tex} carries the cwebmac niceties for prose: \cw{\bs.\char123...\char125}
sets typewriter text (with \cw{\bs\bs}, \cw{\bs\char123}, $\ldots$ for literal
specials), \cw{\bs\bs\char123 id\char125} sets an italic identifier, and the
language tags \cw{\bs CEE/}, \cw{\bs GO/}, \cw{\bs UNIX/}, \cw{\bs TEX/} are
available. \cw{\bs,} is a thin space in math or text. The \.{GWEB} idiom for inline
code, though, remains simply \cw{|...|}.

\section{Customizing the page}
Redefine these in limbo (after the auto-emitted \cw{\bs input gwebmac}):

\bull \cw{\bs def\bs title\char123...\char125} --- the running-head and contents
title (it defaults to the upper-cased file name, and is shown all-caps in the
header);

\bull \cw{\bs pagewidth}, \cw{\bs pageheight} --- the page size;

\bull \cw{\bs def\bs contentspagenumber\char123 n\char125} together with
\cw{\bs pageno} --- to renumber for binding into a larger document;

\bull \cw{\bs Gcodepenalty} (how strongly a code part resists a page break),
\cw{\bs Gcommentkern} (the \cw{//} tightening), and \cw{\bs Gthinmuskip} (the
thin space before \cw{(}, \cw{2mu} by default) are registers you may tune.

\noindent And \cw{\bs datethis} in limbo prints the date and time of the run
beneath section~1, as in \.{CWEB}.

\chapter{Weaving in another language}
\noindent Every fixed phrase \cw{gweave} writes into the woven output --- the
cross-reference notes beneath a code part, the {\it Names of the Sections\/}
heading, the \cw{Section} and \cw{Page} words on the contents page --- goes through
a macro (\cw{\bs GU}, \cw{\bs GA}, \cw{\bs GNused}, \cw{\bs Gsectionword},
\cw{\bs Gpageword}, $\ldots$), never as literal English. Redefine those in an
\cw{\bs input} file and the document is localized to any language; pair the file
with a Unicode-capable engine and suitable fonts, and the code and prose themselves
may be non-ASCII too.

Korean support ships ready-made. Put one line in your limbo,
\begincode
\input kotexgweb
|endcode
weave as usual, and process the result with \cw{luatex} rather than \cw{pdftex}:
\begincode
gweave foo.w
luatex foo.tex
|endcode
\cw{kotexgweb.tex} loads \cw{luatexko} and the Noto Serif/Sans CJK\thinspace KR
fonts, translates \cw{gweave}'s wording into Korean, and supplies a LuaTeX PDF
back end so the blue cross-reference links and the bookmark outline --- with
Korean titles --- keep working; \cw{gweave} itself needs no special flag. The
\cw{hangul} example is a complete web built this way, and \cw{kotexgweb.tex} is
the file to copy as a model for a language of your own.

\chapter{Running the programs}
\begincode
gtangle [-o dir] file[.w] [change[.ch]]
gweave  [-o dir] file[.w] [change[.ch]]
|endcode
The \cw{.w} extension --- and a change file's \cw{.ch} --- may be omitted.
\cw{-o} chooses the output directory. \cw{-version} prints the version and exits.
As each tool runs it prints a one-line banner and then a \cw{*}{\it n\/} progress
marker for every starred section, as \.{CWEB}'s tools do.

\cw{gtangle} always emits \cw{//line} directives in the tangled \GO/ --- as \.{CWEB}'s
\cw{ctangle} always emits \cw{\#line} --- so the compiler, \cw{go vet}, and panic
traces report errors at \cw{.w} positions rather than \cw{.go} ones. \.{GWEB} tangles
its own sources this way too, so editing a \cw{.w} reshuffles those line numbers
in the committed \GO/.

To make a PDF, either go straight through \cw{pdftex}, or through DVI and
\cw{dvipdfmx} for the same hyperlinks:
\begincode
pdftex foo.tex                       # native PDF, with links
tex "\let\pdf+ \input foo.tex"       # or via DVI...
dvipdfmx foo.dvi                     # ...then to PDF, with links
|endcode
Make sure \TeX\ can find \cw{gwebmac.tex} (the installer below puts it on the
search path; otherwise set \cw{TEXINPUTS}).

Both tools report, each with a \cw{file:line} location, an unterminated control
code, a reference to an undefined section, an ambiguous \cw{...} abbreviation, or
a section defined but never used. These are warnings; \cw{gtangle} stops only if
it must actually expand a reference that was never defined.

\chapter{Installing \.{GWEB}}
\noindent Only \cw{gtangle} is committed as \GO/, so it alone installs the \GO/
way:
\begincode
go install github.com/sjnam/gweb/cmd/gtangle@latest
|endcode
\cw{gweave} is generated from its \cw{.w} source rather than committed, so the two
commands are built together from a checkout: \cw{make install-tools} puts both
binaries in your \cw{GOBIN}.

For the full installation --- the commands, \cw{gwebmac.tex} placed in your \TeX\
tree so \cw{pdftex foo.tex} just works, and the man pages --- run
\cw{./install.sh} from a checkout (\cw{./install.sh --uninstall} removes it;
\cw{make install} calls the script). The repository also includes a
VS\thinspace Code syntax highlighter for \cw{.w} files under \cw{editors/vscode},
and this manual is built with \cw{make manual}.

\chapter{The bundled examples}
The \cw{examples/} directory holds six worked webs, from simple to rich; \cw{make
example} tangles and weaves them all, or \cw{make -C examples NAME=wc} just one:

\bull \cw{wc.w} --- a word counter whose tangled output matches the system
\cw{wc}; shows \cw{@f} and a change file (\cw{wc.ch});

\bull \cw{seq.w} --- composable lazy iterators over infinite Fibonacci numbers
(generics, closures, range-over-func);

\bull \cw{squint.w} --- lazy power series as demand-driven channel networks,
after McIlroy's {\it Squinting at Power Series};

\bull \cw{floyd.w} --- Floyd's partition problem, after Knuth's essay {\it Are
Toy Problems Useful?} (meet-in-the-middle search and a \cw{math/big} check);

\bull \cw{sham.w} --- a port of Knuth's Stanford GraphBase demonstration program
\cw{sham}, built on the \cw{go-sgb} module ---\ an external dependency and \cw{goto}
labels;

\bull \cw{hangul.w} --- a short Fibonacci program written in Korean, woven with
\cw{luatex} and \cw{kotexgweb.tex} (see {\it Weaving in another language\/}).

\bull \cw{ugly.w} --- a deliberately messy program designed to test \.{gweave}.
This program implements a 1-indexed Fenwick tree.

\noindent \.{GWEB} also documents itself: \cw{make selfdoc} weaves the system's own
\cw{.w} sources into a single PDF, and \cw{make manual} builds this document.

\vfill
\centerline{\eightrm {\eighttt GWEB} --- a literate programming system for Go.}
\bye
